---
title: "Bacterial Diversity Across Trade Biomes"
author: "Moegamad Uzair Jack"
date: "`r Sys.Date()`"
output: html_document
---

# Loading Packages
```{r}
library(phyloseq)
library(tidyverse)
library(plyr)
library(vegan)
library(ggplot2)
library(grid)
library(RColorBrewer)
library(viridis)
library(pals)
library(MASS)
```

# Set Working Directory
```{r}
setwd("C:/Users/Uzair/Documents/R/Marine_Microbial_Ecology")
```

# Import OTU and Taxonomy Tables
```{r}
otu_TO <- read.table("TO_OTU.txt", header=TRUE)
tax_TO <- read.delim("TO_TAX.txt", header=TRUE)
rownames(tax_TO) <- tax_TO$X
tax_TO$X <- NULL
tax_TO <- as.matrix(tax_TO)
otu_TO <- otu_table(otu_TO, taxa_are_rows = TRUE)
tax_TO <- tax_table(tax_TO)
TO <- merge_phyloseq(otu_TO, tax_TO)
```

# Load and Merge Metadata
```{r}
META2 <- read.delim("META2.txt", header=TRUE)
META <- read.delim("ALL_META.txt", header=TRUE)
META_TO <- merge(META2, META, by="PANGAEA")
META_TO <- column_to_rownames(META_TO, "PANGAEA")
MAP <- sample_data(META_TO)
TO1 <- merge_phyloseq(TO, MAP)
```

# Subset Bacteria in Trades Biome and DCM
```{r}
TO.bacteria <- subset_taxa(TO1, Kingdom == "Bacteria")
TO.trades.bac <- subset_samples(TO.bacteria, Marine_pelagic_biomes_Longhurst_2007 == "Trades")
TO.trades.bac.DCM <- subset_samples(TO.trades.bac, EnvironmentalFeature== "(DCM)")
TO.trades.bac.DCM <- subset_samples(TO.trades.bac.DCM, !is.na(Mean_Salinity_PSU))
```

# Normalize to Relative Abundance
```{r}
TO.trades.bac.DCM.rel <- transform_sample_counts(TO.trades.bac.DCM, function(x) x / sum(x))
TO.trades.bac.DCM.rel.mean <- filter_taxa(TO.trades.bac.DCM.rel, function(x) mean(x) > 0, TRUE)
```

# Alpha Diversity Comparison
```{r}
# Calculate alpha diversity metrics
plot_richness(TO.trades.bac.DCM, x = "Ocean_sea_regions", 
        color = "Ocean_sea_regions", 
        measures = c("Shannon", "Chao1")) +
  theme_bw() +
  theme(
  axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
  axis.text.y = element_text(size = 10),
  axis.title = element_text(size = 12, face = "bold"),
  plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
  plot.subtitle = element_text(size = 12, hjust = 0.5),
  legend.title = element_text(size = 10, face = "bold"),
  panel.grid.minor = element_blank(),
  strip.background = element_rect(fill = "white"),
  strip.text = element_text(face = "bold")
  ) +
  labs(
  title = "Bacterial Alpha Diversity in Trade Wind Biomes",
  subtitle = "Deep Chlorophyll Maximum (DCM) Layer",
  x = "Ocean Region",
  y = "Alpha Diversity Measure"
  )


```

# Ordination: PCoA
```{r}
# Perform PCoA ordination
GP.ord <- ordinate(TO.trades.bac.DCM.rel.mean, method = "PCoA", distance = "bray")

# Extract percent variance explained for axes 1 and 2
pcoa_var <- GP.ord$values$Relative_eig * 100
x_lab <- paste0("PCoA1 [", round(pcoa_var[1], 1), "%]")
y_lab <- paste0("PCoA2 [", round(pcoa_var[2], 1), "%]")

# Create enhanced plot with ellipses and salinity
plot_ordination(TO.trades.bac.DCM.rel.mean, GP.ord, type = "samples", 
     color = "Ocean_sea_regions", shape = "Ocean_sea_regions") +
  geom_point(aes(size = Mean_Salinity_PSU), alpha = 0.8) +
  stat_ellipse(aes(group = Ocean_sea_regions), type = "t", level = 0.95) +
  scale_color_brewer(palette = "Set1", name = "Ocean Region") +
  scale_shape_manual(values = c(16, 17, 15, 18, 19), name = "Ocean Region") +
  scale_size_continuous(name = "Salinity (PSU)") +
  theme_bw() +
  theme(
  axis.text = element_text(size = 12),
  axis.title = element_text(size = 14, face = "bold"),
  plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
  plot.subtitle = element_text(size = 12, hjust = 0.5),
  legend.title = element_text(size = 12, face = "bold"),
  legend.text = element_text(size = 10),
  panel.grid.minor = element_blank()
  ) +
  labs(
  title = "Principal Coordinates Analysis of Bacterial Communities",
  subtitle = "Trade Wind Biomes at Deep Chlorophyll Maximum",
  x = x_lab,
  y = y_lab
  )
```
# Beta Diversity Comparison (Ocean_sea_regions)
```{r}
# Calculate beta diversity using Bray-Curtis dissimilarity
gp_dist <- distance(TO.trades.bac.DCM.rel.mean, method = "bray")
# Extract metadata for PERMANOVA
metadata <- data.frame(sample_data(TO.trades.bac.DCM.rel.mean))
# Perform PERMANOVA
gp_permanova <- adonis2(gp_dist ~ Ocean_sea_regions,
                       data = metadata,
                       permutations = 999)
# Print PERMANOVA results
print(gp_permanova)
``` 


# NMDS Ordination
```{r}
# Convert phyloseq object to matrix for NMDS
otu_matrix <- as(otu_table(TO.trades.bac.DCM.rel.mean), "matrix")
if(taxa_are_rows(TO.trades.bac.DCM.rel.mean)) {
    otu_matrix <- t(otu_matrix)
}

# Perform NMDS ordination
nmds_result <- metaMDS(otu_matrix, distance = "bray", k = 2, trymax = 100, trace = FALSE)

# Create ordination object compatible with phyloseq
nmds_coords <- data.frame(nmds_result$points)
rownames(nmds_coords) <- sample_names(TO.trades.bac.DCM.rel.mean)
ord_obj <- ordinate(TO.trades.bac.DCM.rel.mean, "NMDS", distance = "bray")

# Create NMDS plot with ellipses and salinity
plot_ordination(TO.trades.bac.DCM.rel.mean, ord_obj, type = "samples",
                color = "Ocean_sea_regions", shape = "Ocean_sea_regions") +
  geom_point(aes(size = Mean_Salinity_PSU), alpha = 0.8) +
  stat_ellipse(aes(group = Ocean_sea_regions), type = "t", level = 0.95) +
  scale_color_brewer(palette = "Set1", name = "Ocean Region") +
  scale_shape_manual(values = c(16, 17, 15, 18, 19), name = "Ocean Region") +
  scale_size_continuous(name = "Salinity (PSU)") +
  theme_bw() +
  theme(
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 10),
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "NMDS Ordination of Bacterial Communities",
    subtitle = "Trade Wind Biomes at Deep Chlorophyll Maximum",
    x = x_lab,
    y = y_lab
  )
```

# Environmental Fitting (NMDS + envfit)
```{r}
# NMDS ordination
psv <- data.frame(sample_data(TO.trades.bac.DCM.rel.mean))
psotu2veg <- function(physeq) {
  OTU <- otu_table(physeq)
  if (taxa_are_rows(OTU)) {
    OTU <- t(OTU)
  }
  return(as(OTU, "matrix"))
}
ps.otu.veg <- psotu2veg(TO.trades.bac.DCM.rel.mean)
vare.mds <- metaMDS(ps.otu.veg, trace = FALSE)

# Environmental fitting
ef <- envfit(vare.mds, psv, permu = 999, na.rm = TRUE)

# Extract NMDS coordinates for samples
nmds_scores <- as.data.frame(scores(vare.mds, display = "sites"))
nmds_scores$Mean_Salinity_PSU <- sample_data(TO.trades.bac.DCM.rel.mean)$Mean_Salinity_PSU
nmds_scores$Ocean_sea_regions <- sample_data(TO.trades.bac.DCM.rel.mean)$Ocean_sea_regions

# Stress value for subtitle
nmds_stress <- round(vare.mds$stress, 3)

# Significant vectors only
sig_vec <- ef$vectors$pvals < 0.05
vecs <- as.data.frame(ef$vectors$arrows[sig_vec, , drop = FALSE])
vecs$var <- rownames(vecs)
arrow_mult <- 1.5
vecs$NMDS1 <- vecs$NMDS1 * arrow_mult
vecs$NMDS2 <- vecs$NMDS2 * arrow_mult

# NMDS plot with environmental vectors
ggplot(nmds_scores, aes(x = NMDS1, y = NMDS2, color = Mean_Salinity_PSU, shape = Ocean_sea_regions)) +
  geom_point(size = 3, alpha = 0.8, stroke = 0.2) +
  scale_color_viridis(option = "plasma", name = "Salinity (PSU)", direction = -1) +
  scale_shape_manual(values = c(16, 17, 15, 18, 19), name = "Ocean Region") +
  geom_segment(
    data = vecs,
    aes(x = 0, y = 0, xend = NMDS1, yend = NMDS2),
    arrow = arrow(length = unit(0.35, "cm")), color = "navy", size = 1.1, inherit.aes = FALSE
  ) +
  geom_text(
    data = vecs,
    aes(x = NMDS1 * 1.12, y = NMDS2 * 1.12, label = var),
    color = "navy", size = 3, fontface = "bold", hjust = 0.5, inherit.aes = FALSE
  ) +
  theme_classic(base_size = 9) +
  labs(
    title = "NMDS Ordination with Significant Environmental Vectors",
    subtitle = paste0("Trade Wind Biomes at DCM (Stress = ", nmds_stress, ")"),
    x = "NMDS1",
    y = "NMDS2",
    caption = "Arrows indicate significant (p < 0.05) environmental variables."
  ) +
  theme(
    plot.title = element_text(size = 9, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 8, hjust = 0.5),
    axis.title = element_text(size = 9, face = "bold"),
    axis.text = element_text(size = 8),
    legend.title = element_text(size = 9, face = "bold"),
    legend.text = element_text(size = 8),
    legend.position = "right"
  )

```

# Barplot of Dominant Bacterial Phyla by Ocean
```{r}
# First identify samples that are truly DCM and exclude TARA_100_MES_0.22-3
truly_dcm <- subset_samples(TO.trades.bac.DCM.rel.mean, EnvironmentalFeature == "(DCM)" & 
              !Sample_label %in% c("TARA_100_MES_0.22-3"))

# Aggregate at Phylum level and get top 10 phyla using the filtered dataset
PhylumGlommed <- tax_glom(truly_dcm, "Phylum")
top_phyla <- names(sort(taxa_sums(PhylumGlommed), decreasing = TRUE)[1:15])
PhylumGlommed_filtered <- prune_taxa(top_phyla, PhylumGlommed)

# Create the plot with the filtered dataset
ggplot(data = psmelt(PhylumGlommed_filtered), 
  aes(x = Sample_label, y = Abundance, fill = Phylum)) +
  geom_bar(stat = "identity", position = "stack", width = 0.8) +
  facet_grid(~Ocean_sea_regions, scales = "free_x", space = "free") +
  scale_fill_manual(values = colorRampPalette(brewer.pal(12, "Paired"))(15)) +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_bw() +
  theme(
  axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
  axis.text.y = element_text(size = 10),
  axis.title = element_text(size = 12, face = "bold"),
  plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
  plot.subtitle = element_text(size = 12, hjust = 0.5),
  legend.title = element_text(size = 10, face = "bold"),
  legend.text = element_text(size = 9),
  legend.position = "right",
  panel.grid.minor = element_blank(),
  strip.text = element_text(size = 10, face = "bold"),
  strip.background = element_rect(fill = "white")
  ) +
  labs(
  title = "Bacterial Phylum Composition Across Ocean Regions",
  subtitle = "Top 15 Most Abundant Phyla in Trade Wind Biomes at DCM",
  x = "Sample",
  y = "Relative Abundance"
  )
```

# Save Plots
```{r}
# Create directory for plots if it doesn't exist
dir.create("plots", showWarnings = FALSE)

# Alpha diversity plot
ggsave("plots/alpha_diversity.png", 
  plot = last_plot(), 
  width = 12, 
  height = 8, 
  dpi = 300)

# PCoA plot
pcoa_plot <- plot_ordination(TO.trades.bac.DCM.rel.mean, GP.ord, type = "samples", 
        color = "Ocean_sea_regions", shape = "Ocean_sea_regions")
ggsave("plots/pcoa_plot.png", 
  plot = pcoa_plot, 
  width = 10, 
  height = 8, 
  dpi = 300)

# NMDS plot
nmds_plot <- plot_ordination(TO.trades.bac.DCM.rel.mean, ord_obj, type = "samples",
        color = "Ocean_sea_regions", shape = "Ocean_sea_regions")
ggsave("plots/nmds_plot.png", 
  plot = nmds_plot, 
  width = 10, 
  height = 8, 
  dpi = 300)

# Environmental fitting plot
ggsave("plots/envfit_plot.png", 
  plot = last_plot(), 
  width = 10, 
  height = 8, 
  dpi = 300)

# Phylum composition barplot
ggsave("plots/phylum_barplot.png", 
  plot = last_plot(), 
  width = 15, 
  height = 8, 
  dpi = 300)
```
