---
title: "Bacterial Communities in Trades Biomes"
author: "Moegamad Uzair Jack"
date: "`r Sys.Date()`"
output: html_document
---

# Load Packages
```{r}
library(phyloseq)
library(tidyverse)
library(vegan)
library(RColorBrewer)
library(pals)
library(MASS)

```

# Set Working Directory
```{r}
setwd("C:/Users/Uzair/Documents/R/Marine_Microbial_Ecology")
```

# Import OTU and Taxonomy Tables
```{r}
otu_data <- read.table("TO_OTU.txt", header = TRUE)
tax_data <- read.delim("TO_TAX.txt", header = TRUE)
tax_data <- column_to_rownames(tax_data, "X")
tax_data <- as.matrix(tax_data)

otu_table_obj <- otu_table(otu_data, taxa_are_rows = TRUE)
tax_table_obj <- tax_table(tax_data)
phylo_obj <- merge_phyloseq(otu_table_obj, tax_table_obj)

```

# Load and Merge Metadata
```{r}
meta2_data <- read.delim("META2.txt", header = TRUE)
meta_data <- read.delim("ALL_META.txt", header = TRUE)
meta_combined <- merge(meta2_data, meta_data, by = "PANGAEA")
meta_combined <- column_to_rownames(meta_combined, "PANGAEA")

sample_map <- sample_data(meta_combined)
phylo_complete <- merge_phyloseq(phylo_obj, sample_map)

```

# Subset: Bacteria Fraction in Trades Biomes
```{r}
bacteria_data <- subset_taxa(phylo_complete, Kingdom == "Bacteria")
trades_bacteria <- subset_samples(
  bacteria_data, 
  Marine_pelagic_biomes_Longhurst_2007 == "Trades"
)

```

# Normalise to Relative Abundance
```{r}
bacteria_data <- transform_sample_counts(trades_bacteria, function(x) x / sum(x))
TO.trades.bac.rel.mean <- filter_taxa(bacteria_data, function(x) mean(x) > 0, TRUE)

```
# Alpha Diversity  by Ocean Region
```{r}
# Calculate alpha diversity metrics
alpha_div <- plot_richness(
  TO.trades.bac.rel.mean,
  x = "Ocean_sea_regions",
  measures = c("Shannon", "Chao1")
)
alpha_data <- alpha_div$data

# Create enhanced plot using ggplot2
ggplot(alpha_data, aes(x = Ocean_sea_regions, y = value, fill = Ocean_sea_regions)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21) +
  geom_jitter(width = 0.2, alpha = 0.5, size = 2) +
  facet_wrap(~variable, scales = "free_y", nrow = 2) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Alpha Diversity of Bacterial Communities",
    subtitle = "Trades Biomes across Ocean Regions",
    x = "Ocean Regions",
    y = "Diversity Index Value",
    fill = "Ocean Regions"
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 12, family = "Arial"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title = element_text(face = "bold"),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 12, face = "italic"),
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold"),
    legend.position = "right"
  )

# Ordination (PCoA)
```{r}
# Perform PCoA ordination
ord_trades <- ordinate(TO.trades.bac.rel.mean, method = "PCoA", distance = "bray")

# Create enhanced plot
plot_ordination(TO.trades.bac.rel.mean, ord_trades, type = "samples", color = "Ocean_sea_regions") +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse(type = "norm", linetype = 2, aes(color = Ocean_sea_regions)) +
  scale_color_brewer(palette = "Set2") +
  labs(
    title = "Principal Coordinates Analysis (PCoA)",
    subtitle = "Bacterial Communities in Trades Biomes",
    x = paste0("PCoA1 [", round(100 * ord_trades$values$Relative_eig[1], 1), "%]"),
    y = paste0("PCoA2 [", round(100 * ord_trades$values$Relative_eig[2], 1), "%]"),
    color = "Ocean Regions"
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 12, family = "Arial"),
    axis.text = element_text(size = 10),
    legend.position = "right",
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 12, face = "italic"),
    panel.grid.minor = element_blank()
  )

```



# Environmental Fitting with NMDS
```{r NMDS_and_envfit_plots, message=FALSE, warning=FALSE}
# This ordination shows how much each sample differs from the other composition wise. When we superimpose
#vectors on it this will tell us how these metadata are positively (pointing towards the sample) or negatively 
# (pointing directly away from a sample dot) influencing those samples (dots)

psv = data.frame(sample_data(TO.trades.bac.rel.mean))

# convert the sample_data() within a phyloseq object to a vegan compatible data object
pssd2veg <- function(physeq) {
  sd <- sample_data(physeq)
  return(as(sd,"data.frame"))
}

# convert the otu_table() within a phyloseq object to a vegan compatible data object
psotu2veg <- function(physeq) {
  OTU <- otu_table(physeq)
  if (taxa_are_rows(OTU)) {
    OTU <- t(OTU)
  }
  return(as(OTU, "matrix"))
}

ps.sd.veg <- pssd2veg(TO.trades.bac.rel.mean)
ps.otu.veg <- psotu2veg(TO.trades.bac.rel.mean)

vare.dis <- vegdist(ps.otu.veg, method="bray")
vare.mds <- metaMDS(ps.otu.veg, trace = FALSE)
vare.mds

data.scores = as.data.frame(scores(vare.mds))

ef <- envfit(vare.mds, psv, permu = 999, na.rm=TRUE)
ef
plot(vare.mds, display = "sites")
plot(ef, p.max = 0.001)
ordiplot(vare.mds, display = 'sites', type = 'n')
orditorp(vare.mds, display = 'sites')

# Create enhanced ordination plots with environmental variables
ord_plot <- plot_ordination(TO.trades.bac.rel.mean, ord_trades, type="samples", color="Ocean_sea_regions")
ord_plot_depth <- plot_ordination(TO.trades.bac.rel.mean, ord_trades, type="samples", color="Mean_depth")

plot(ord_plot)
plot(ord_plot_depth)

```
# Relative Abundance of Bacterial Classes
class_glommed <- tax_glom(trades_bacteria, taxrank = "Class")
# Agglomerate taxa at Class level
ClassGlommed <- tax_glom(trades_bacteria, taxrank = "Class")

# Ensure Sample_label exists in sample data
if (!"Sample_label" %in% colnames(sample_data(ClassGlommed))) {
  sample_data(ClassGlommed)$Sample_label <- rownames(sample_data(ClassGlommed))
}

# Calculate relative abundance
ClassGlommed <- transform_sample_counts(ClassGlommed, function(x) x / sum(x))

# Get number of unique classes for color palette
n_classes <- length(unique(tax_table(ClassGlommed)[, "Class"]))
palette_expanded <- colorRampPalette(brewer.pal(12, "Paired"))(n_classes)

# Create enhanced plot
plot_bar(ClassGlommed, "Sample_label", fill = "Class") +
  facet_grid(~Ocean_sea_regions, scales = "free_x", space = "free") +
  scale_fill_manual(values = palette_expanded) +
  labs(
    title = "Relative Abundance of Bacterial Classes",
    subtitle = "Distribution across Ocean Regions in Trades Biomes",
    x = "Samples",
    y = "Relative Abundance",
    fill = "Bacterial Class"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold", size = 12),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 12, face = "italic"),
    legend.position = "right",
    legend.text = element_text(size = 8),
    axis.title = element_text(face = "bold")
  )
    axis.title = element_text(face = "bold")
  )
    axis.title = element_text(face = "bold")
  )
```

