---
title: "NPO Sample Analysis: Understanding Differences in Bacterial Communities"
author: "Moegamad Uzair Jack"
date: "`r Sys.Date()`"
output:
  html_document: default
  word_document: default
---

# Loading Packages

```{r setup, message=FALSE, warning=FALSE}
library(phyloseq)
library(tidyverse)
library(vegan)
library(ggplot2)
library(RColorBrewer)
library(viridis)
library(reshape2)
library(scales)
```

# Set Working Directory and Load Data

```{r}
setwd("C:/Users/Uzair/Documents/R/Marine_Microbial_Ecology")

# Import OTU and Taxonomy Tables
otu_TO <- read.table("TO_OTU.txt", header=TRUE)
tax_TO <- read.delim("TO_TAX.txt", header=TRUE)
rownames(tax_TO) <- tax_TO$X
tax_TO$X <- NULL
tax_TO <- as.matrix(tax_TO)
otu_TO <- otu_table(otu_TO, taxa_are_rows = TRUE)
tax_TO <- tax_table(tax_TO)
TO <- merge_phyloseq(otu_TO, tax_TO)

# Load and Merge Metadata
META2 <- read.delim("META2.txt", header=TRUE)
META <- read.delim("ALL_META.txt", header=TRUE)
META_TO <- merge(META2, META, by="PANGAEA")
META_TO <- column_to_rownames(META_TO, "PANGAEA")
MAP <- sample_data(META_TO)
TO1 <- merge_phyloseq(TO, MAP)

# Subset Bacteria in Trades Biome and DCM
TO.bacteria <- subset_taxa(TO1, Kingdom == "Bacteria")
TO.trades.bac <- subset_samples(TO.bacteria, Marine_pelagic_biomes_Longhurst_2007 == "Trades")
TO.trades.bac.DCM <- subset_samples(TO.trades.bac, EnvironmentalFeature== "(DCM)")
TO.trades.bac.DCM <- subset_samples(TO.trades.bac.DCM, !is.na(Mean_Salinity_PSU))

# Normalize to Relative Abundance
TO.trades.bac.DCM.rel <- transform_sample_counts(TO.trades.bac.DCM, function(x) x / sum(x))
TO.trades.bac.DCM.rel.mean <- filter_taxa(TO.trades.bac.DCM.rel, function(x) mean(x) > 0, TRUE)
```

# Check Available Samples and Extract NPO

```{r}
# Check what ocean regions are available
print("Available Ocean Regions:")
print(table(sample_data(TO.trades.bac.DCM.rel.mean)$Ocean_sea_regions))

# Extract NPO samples (if they exist)
npo_check <- sample_data(TO.trades.bac.DCM.rel.mean)$Ocean_sea_regions == "NPO"
print(paste("Number of NPO samples found:", sum(npo_check, na.rm = TRUE)))

if(sum(npo_check, na.rm = TRUE) > 0) {
  npo_samples <- subset_samples(TO.trades.bac.DCM.rel.mean, Ocean_sea_regions == "NPO")
  npo_meta <- data.frame(sample_data(npo_samples))
  
  print("NPO Sample Information:")
  print(npo_meta[, c("Sample_label", "Ocean_sea_regions")])
  print(paste("Number of NPO samples:", nsamples(npo_samples)))
} else {
  print("No NPO samples found. Available regions are:")
  print(unique(sample_data(TO.trades.bac.DCM.rel.mean)$Ocean_sea_regions))
  
  # You might need to adjust the region name or choose a different region
  # For example, if the region is named differently:
  # npo_samples <- subset_samples(TO.trades.bac.DCM.rel.mean, Ocean_sea_regions == "North_Pacific")
}
```

# Extract NPO Samples for Analysis

```{r}
# Extract only NPO samples
npo_samples <- subset_samples(TO.trades.bac.DCM.rel.mean, Ocean_sea_regions == "(NPO)")
npo_meta <- data.frame(sample_data(npo_samples))

print("NPO Sample Information:")
print(npo_meta[, c("Sample_label", "Ocean_sea_regions")])
print(paste("Number of NPO samples:", nsamples(npo_samples)))
```

# Environmental Variable Comparison

```{r}
# Compare environmental variables between NPO samples
env_vars_compare <- c("Mean_Lat", "Mean_Long", "Mean_Depth_m", "Mean_Temperature_degC",
                     "Mean_Salinity_PSU", "Mean_Oxygen_umolkg", "Mean_Nitrates_umolL")

npo_env <- npo_meta[, env_vars_compare]
rownames(npo_env) <- npo_meta$Sample_label

print("NPO Environmental Conditions:")
print(npo_env)

# Calculate differences between samples
if(nrow(npo_env) == 2) {
  env_differences <- abs(npo_env[1,] - npo_env[2,])
  print("Environmental Differences Between NPO Samples:")
  print(env_differences)
}
```

# Visualize Environmental Differences

```{r}
# Prepare data for visualization
npo_env$Sample <- rownames(npo_env)
npo_long <- melt(npo_env, id.vars = "Sample")

# Plot environmental differences
ggplot(npo_long, aes(x = variable, y = value, fill = Sample)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.8) +
  facet_wrap(~variable, scales = "free", ncol = 3) +
  theme_bw() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    strip.text = element_text(face = "bold"),
    legend.position = "bottom"
  ) +
  labs(
    title = "Environmental Differences Between NPO Samples",
    subtitle = "Comparison of Key Environmental Variables",
    x = "",
    y = "Value"
  ) +
  scale_fill_brewer(palette = "Set2")

ggsave("plots/npo_environmental_comparison.png", 
       width = 12, height = 8, dpi = 300)
```

# Alpha Diversity Comparison

```{r}
# First, extract NPO samples from the COUNT data (not relative abundance)
npo_samples_counts <- subset_samples(TO.trades.bac.DCM, Ocean_sea_regions == "(NPO)")

# Calculate alpha diversity for NPO samples using COUNT data
npo_alpha <- estimate_richness(npo_samples_counts, measures = c("Shannon", "Chao1", "Simpson"))

npo_alpha$Sample <- c("TARA_137_DCM_0.22-3", "TARA_138_DCM_0.22-3")

print("Alpha Diversity Comparison:")
print(npo_alpha)


# Visualize alpha diversity differences
npo_alpha_long <- melt(npo_alpha, id.vars = "Sample")

ggplot(npo_alpha_long, aes(x = Sample, y = value, fill = Sample)) +
    geom_bar(stat = "identity", alpha = 0.8) +
    facet_wrap(~variable, scales = "free") +
    theme_bw() +
    theme(
        axis.text.x = element_text(hjust = 0.5, size = 8),
        strip.text = element_text(face = "bold"),
        legend.position = "none"
    ) +
    labs(
        title = "Alpha Diversity Comparison: NPO Samples",
        x = "Sample",
        y = "Diversity Index Value"
    ) +
    scale_fill_brewer(palette = "Set2")

ggsave("plots/npo_alpha_diversity.png", 
             width = 10, height = 6, dpi = 300)


## NB: TARA_B100001964	TARA_137_DCM_0.22-3	(NPO)		
##    TARA_B100001996	TARA_138_DCM_0.22-3	(NPO)
```

# Beta Diversity Analysis

```{r}
# Calculate Bray-Curtis dissimilarity between NPO samples
npo_dist <- distance(npo_samples, method = "bray")

print("Bray-Curtis Dissimilarity Between NPO Samples:")
print(as.matrix(npo_dist))

# If you have more than 2 NPO samples, create a heatmap
if(nsamples(npo_samples) > 2) {
  dist_matrix <- as.matrix(npo_dist)
  dist_melt <- melt(dist_matrix)
  
  ggplot(dist_melt, aes(Var1, Var2, fill = value)) +
    geom_tile() +
    scale_fill_viridis_c(name = "Bray-Curtis\nDissimilarity") +
    theme_minimal() +
    labs(
      title = "Bray-Curtis Dissimilarity Matrix: NPO Samples",
      x = "Sample",
      y = "Sample"
    )
}
```

# Bacterial Community Composition Analysis

```{r}
# Compare at Class level
# Agglomerate at Class level
npo_class <- tax_glom(npo_samples, "Class")

# Get top 15 Classes
top_class <- names(sort(taxa_sums(npo_class), decreasing = TRUE)[1:15])
npo_class_filtered <- prune_taxa(top_class, npo_class)

# Plot class differences
ggplot(data = psmelt(npo_class_filtered), 
    aes(x = Sample_label, y = Abundance, fill = Class)) +
  geom_bar(stat = "identity", position = "stack", width = 0.8) +
  facet_grid(~Ocean_sea_regions, scales = "free_x", space = "free") +
  scale_fill_manual(values = colorRampPalette(brewer.pal(12, "Paired"))(15)) +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_bw() +
  theme(
    axis.text.x = element_text(hjust = 0.5, size = 8),,
    axis.text.y = element_text(size = 10),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    legend.position = "right",
    panel.grid.minor = element_blank(),
    strip.text = element_text(size = 10, face = "bold"),
    strip.background = element_rect(fill = "white")
  ) +
  labs(
    title = "Bacterial Class Composition in NPO Samples",
    subtitle = "Top 15 Most Abundant Classes at DCM",
    x = "Sample",
    y = "Relative Abundance"
  )

ggsave("plots/npo_class_composition.png", 
    width = 14, height = 8, dpi = 300)
```

# Geographic Analysis

```{r}
# Analyze geographic and sampling differences
npo_geo <- npo_meta[, c("Mean_Lat", "Mean_Long", "Sample_label")]

print("NPO Sample Geographic Locations:")
print(npo_geo)

# Calculate geographic distance if 2 samples
if(nrow(npo_geo) == 2) {
  lat_diff <- abs(npo_geo$Mean_Lat[1] - npo_geo$Mean_Lat[2])
  long_diff <- abs(npo_geo$Mean_Long[1] - npo_geo$Mean_Long[2])
  
  # Approximate distance calculation (rough estimate)
  geo_distance <- sqrt(lat_diff^2 + long_diff^2) * 111 # Convert to km
  
  print(paste("Latitude difference:", round(lat_diff, 2), "degrees"))
  print(paste("Longitude difference:", round(long_diff, 2), "degrees"))
  print(paste("Approximate distance:", round(geo_distance, 0), "km"))
}

# Plot sample locations
# Add geographic context
# Load required packages for mapping
library(maps)
library(ggspatial)

# Get world map data
world_map <- map_data("world")

ggplot(npo_geo, aes(x = Mean_Long, y = Mean_Lat, color = Sample_label)) +
  geom_polygon(data = world_map, aes(x = long, y = lat, group = group),
               fill = "lightgray", color = "white") +
  geom_point(size = 4) +
  geom_text(aes(label = Sample_label), vjust = -1) +
  annotation_scale(location = "bl") +
  annotation_north_arrow(location = "tr",
                        style = north_arrow_fancy_orienteering) +
  coord_fixed(1.3) +
  theme_minimal()


ggsave("plots/npo_geographic_locations.png", 
       width = 10, height = 6, dpi = 300)
```

# Summary and Conclusions

```{r}
cat("
## Summary of NPO Sample Analysis

### Key Findings:
1. Environmental Differences: 
   - [Describe the main environmental differences found]
   
2. Community Composition:
   - [Summarize the main bacterial community differences]
   
3. Alpha Diversity:
   - [Compare diversity patterns between samples]
   
4. Geographic Factors:
   - [Discuss geographic distance and its potential influence]

### Potential Explanations:
- Environmental gradients (salinity, temperature, nutrients)
- Geographic distance effects
- Local oceanographic conditions
- Different water masses

### Next Steps:
- Investigate specific environmental drivers
- Compare with other regions
- Consider temporal variation if applicable
")
```

# Create Summary Report

```{r}
# Create a summary table of all findings
env_diff_summary <- data.frame(
  Variable = names(env_differences),
  Absolute_Difference = as.numeric(env_differences),
  Relative_Difference = as.numeric(env_differences) / colMeans(npo_env[,1:7])
)

summary_report <- list(
  "Sample_Count" = nsamples(npo_samples),
  "Geographic_Distance_km" = if(exists("geo_distance")) round(geo_distance, 0) else "Not calculated",
  "Community_Dissimilarity" = round(as.numeric(npo_dist), 3),
  "Top_Environmental_Driver" = env_diff_summary$Variable[which.max(abs(env_diff_summary$Relative_Difference))],
  "Diversity_Difference" = round(abs(npo_alpha$Shannon[1] - npo_alpha$Shannon[2]), 3)
)

print("NPO Analysis Summary:")
print(summary_report)
```
# Oxygen and Nutrient Analysis
```{r}
# Extract oxygen and nutrient data
nutrient_data <- data.frame(
  Sample = rownames(npo_env),
  Oxygen = npo_env$Mean_Oxygen_umolkg,
  Nitrates = npo_env$Mean_Nitrates_umolL
)

# Modify plot_shared_theme
plot_shared_theme <- theme_bw() +
  theme(
    axis.text.x = element_text(angle = 15, hjust = 0.8),  # Better angle for readability
    plot.title = element_text(size = 9, face = "bold"), # Larger title
    panel.grid.minor = element_blank(),
    plot.subtitle = element_text(size = 10, color = "darkgray")  # Add subtitle for context
  )

# Add reference ranges to plots
p1 <- p1 +
  geom_hline(yintercept = c(100, 150), linetype = "dashed", color = "gray50") +
  annotate("text", x = 1.5, y = 155, label = "Typical range", size = 3)

p2 <- p2 +
  geom_hline(yintercept = c(5, 15), linetype = "dashed", color = "gray50") +
  annotate("text", x = 1.5, y = 16, label = "Typical range", size = 3)

# Oxygen plot
p1 <- ggplot(nutrient_data, aes(x = Sample, y = Oxygen, fill = Sample)) +
  geom_bar(stat = "identity", width = 0.7) +
  plot_shared_theme +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Oxygen Levels",
    y = "Oxygen (μmol/kg)"
  ) +
  theme(legend.position = "none")

# Nitrate plot
p2 <- ggplot(nutrient_data, aes(x = Sample, y = Nitrates, fill = Sample)) +
  geom_bar(stat = "identity", width = 0.7) +
  plot_shared_theme +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Nitrate Levels",
    y = "Nitrates (μmol/L)"
  ) +
  theme(legend.position = "none")

# Combine plots
library(gridExtra)
grid.arrange(p1, p2, ncol = 2,
             top = "Oxygen and Nutrient Levels at NPO Sample Sites")

# Save combined plot
ggsave("plots/npo_nutrients_separate.png", 
       arrangeGrob(p1, p2, ncol = 2),
       width = 12, height = 5, dpi = 300)
```
# Add nutrient analysis and correlations
```{r}
# Calculate basic statistics
nutrient_stats <- data.frame(
  Parameter = c("Oxygen", "Nitrates"),
  Mean = c(mean(nutrient_data$Oxygen), mean(nutrient_data$Nitrates)),
  SD = c(sd(nutrient_data$Oxygen), sd(nutrient_data$Nitrates)),
  Difference = c(diff(nutrient_data$Oxygen), diff(nutrient_data$Nitrates))
)

# Add correlation with bacterial classes
tax_nutrients <- data.frame(
  Oxygen = rep(nutrient_data$Oxygen, each = ntaxa(npo_class_filtered)),
  Nitrates = rep(nutrient_data$Nitrates, each = ntaxa(npo_class_filtered)),
  Class = rep(top_class, times = 2),
  Abundance = psmelt(npo_class_filtered)$Abundance
)

print("Nutrient Statistics:")
print(nutrient_stats)

# Compare to literature DCM values
cat("\nLiterature Context:\n")
cat("- DCM oxygen ranges in tropical Pacific: 
     * 100-150 μmol/kg (Paulmier & Ruiz-Pino, 2009 - Progress in Oceanography)
     * 50-200 μmol/kg in Costa Rica Dome (Fiedler & Talley, 2006 - Progress in Oceanography)\n")
cat("- DCM nitrate ranges: 
     * 5-15 μmol/L in tropical Pacific DCM (Johnson et al., 2010 - Nature Geoscience)
     * 10-20 μmol/L in Costa Rica Dome upwelling (Pennington et al., 2006 - Progress in Oceanography)\n")

# Add references section
cat("\nReferences:\n")
cat("1. Paulmier, A., & Ruiz-Pino, D. (2009). Oxygen minimum zones (OMZs) in the modern ocean. Progress in Oceanography, 80(3-4), 113-128.\n")
cat("2. Fiedler, P. C., & Talley, L. D. (2006). Hydrography of the eastern tropical Pacific: A review. Progress in Oceanography, 69(2-4), 143-180.\n")
cat("3. Johnson, K. S., et al. (2010). Ocean nutrient ratios governed by plankton biogeography. Nature Geoscience, 3(9), 625-629.\n")
cat("4. Pennington, J. T., et al. (2006). Primary production in the eastern tropical Pacific: A review. Progress in Oceanography, 69(2-4), 285-317.\n")
```
```
